Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-24 main.tex
typo

First-order predicate logic is one of the cornerstones of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer science. Since
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on its own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
that has been made in proof theory in general and in the area of
automated and interactive theorem provers in
particular\cite{DSAV:15lean,paulson:isabelle}\todo{find
references}\juihsuan{done}, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

  Conversely, assume we have a decomposed derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  Then we can transform $\Deri_1$ into a recified form
  $\rectif\Deri_1$, proving $\rectif A'$. By
  Theorem~\ref{thm:MLS1->fonet}, the linked fograph
  $\fographof{\rectif\Deri_1}=\tuple{\fographof{\rectif
      A'},\linkingof{\rectif\Deri_1}}$ is a fonet.  Then, by
  Lemma~\ref{lem:cw->rectif}, there is a rectified derivation
  $\vlderivation{
    \vlde{\set{\wrD,\rectif\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif{\Deri_2}}{\rectif
      A}{ \vlhy{\rectif{A'}}}}$ whose induced map
  $\mapof{\rectif{\Deri_2}}\colon\graphof{\rectif{A'}}\to\graphof{\rectif
    A}$ is the same a the induced map
  $\mapof{\Deri_2}\colon\graphof{A'}\to\graphof{A}$ of $\Deri_2$. By
  Lemma~\ref{lem:cw->skew}, this map is a skew bifibration. Hence, we have a combinatorial proof \lutz{shit, something's wrong...}

  %% Let $\downsmash{\vlderivation{
  %%   \vlde{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
  %%     \vlhy{A}}}}$ be a derivation. Then there is a recified derivation
  %% $\vlderivation{
  %%   \vlde{\set{\wrD,\rectif\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{\rectif
  %%     B}{ \vlhy{\rectif A}}}$, such that the induced maps
  %% $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ and
  %% $\mapof{\rectif\Deri}\colon\graphof{\rectif A}\to\graphof{\rectif
  %%   B}$ are equal up to a variable renaming of the vertex labels.

  %% for some $A'\fequ C\sigma$ where $C$ is a formula with $\graphof
  %% C=\gC$ and $\sigma$ is a variable renaming substitution.  Conversely,
  %% whenever we have a derivation as in~\eqref{eq:decom} above, then
  %% there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such
  %% that $\gC=\graphof{\rectif{A'}}$.

@@

>>>>

First-order predicate logic is one of the cornerstones of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer science. Since
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on their own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
that has been made in proof theory in general and in the area of
automated and interactive theorem provers in
particular\cite{DSAV:15lean,paulson:isabelle}\todo{find
references}\juihsuan{done}, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

  Conversely, assume we have a decomposed derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  Then we can transform $\Deri_1$ into a recified form
  $\rectif\Deri_1$, proving $\rectif A'$. By
  Theorem~\ref{thm:MLS1->fonet}, the linked fograph
  $\fographof{\rectif\Deri_1}=\tuple{\fographof{\rectif
      A'},\linkingof{\rectif\Deri_1}}$ is a fonet.  Then, by
  Lemma~\ref{lem:cw->rectif}, there is a rectified derivation
  $\vlderivation{
    \vlde{\set{\wrD,\rectif\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif{\Deri_2}}{\rectif
      A}{ \vlhy{\rectif{A'}}}}$ whose induced map
  $\mapof{\rectif{\Deri_2}}\colon\graphof{\rectif{A'}}\to\graphof{\rectif
    A}$ is the same a the induced map
  $\mapof{\Deri_2}\colon\graphof{A'}\to\graphof{A}$ of $\Deri_2$. By
  Lemma~\ref{lem:cw->skew}, this map is a skew bifibration. Hence, we
  have a combinatorial proof $\phi\colon\gC\to\graphof{A}$ with
  $\gC=\graphof{\rectif{A'}}$.
  \lutz{shit, something's wrong...}
  %% Let $\downsmash{\vlderivation{
  %%   \vlde{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
  %%     \vlhy{A}}}}$ be a derivation. Then there is a recified derivation
  %% $\vlderivation{
  %%   \vlde{\set{\wrD,\rectif\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{\rectif
  %%     B}{ \vlhy{\rectif A}}}$, such that the induced maps
  %% $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ and
  %% $\mapof{\rectif\Deri}\colon\graphof{\rectif A}\to\graphof{\rectif
  %%   B}$ are equal up to a variable renaming of the vertex labels.
  %% for some $A'\fequ C\sigma$ where $C$ is a formula with $\graphof
  %% C=\gC$ and $\sigma$ is a variable renaming substitution.  Conversely,
  %% whenever we have a derivation as in~\eqref{eq:decom} above, then
  %% there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such
  %% that $\gC=\graphof{\rectif{A'}}$.
\end{proof}

Note that Theorem~\ref{thm:CP-DI} shows at the same time soundness, completeness, and full completeness, as
\begin{enumerate}
\item every proof in $\FOKS$ can be translated into a combinatorial proof, and
\item every combinatorial proof is the image of a $\FOKS$-proof under that translation.
\end{enumerate}

@@

In this paper we have uncovered a close correspondence between
first-order combinatorial proofs and decomposed deep inference
derivations of system $\FOKS$, and we have shown that every proof in
$\FOKS$ has such a decomposed form.

The most surprising discovery for us was that all technical
difficulties in our work could be reduced (in a non-trivial way) to
the propositional setting.

The obvious next step in our research is to investigate proof
composition and normalisation of first-order combinatorial
proofs. Even in the propositional setting, the normalization of
combinatorial proofs is underdevelopped. There exist two different
procedures for cut elimination~\cite{hughes:invar,str:FSCD17}, but
both have their insufficiencies, and there is no general theory.

\lutz{do we want/can say more here?}
